---
title: "final report"
authors: 'Anders Mellbye, Juan Pablo López Cervantes, Guri Kringeland '
date: "2025-09-11"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

tinytex::install_tinytex()

##INTRODUCTION

Source: This data set is mainly from Cata et al. ’Blood Storage Duration
and Biochemical Recurrence of Cancer after Radical Prostate- ctomy’.
Mayo Clin Proc 2011; 86(2): 120-127. A clean version is taken from R
package medicaldata.

From the description on package website: “This data set contains data on
316 men who had undergone radical prostatectomy and received transfusion
during or within 30 days of the surgical procedure and had available
prostate serum antigen (PSA) follow-up data. The main exposure of
interest was RBC storage duration group.”

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Tidying the data The datasett contained oirginally 672 rows for 316
IDs and 21 variables.

First we deleted 40 duplicates. Through the pivot wider task (
volumemasurement and .value-\> PVol and TVol). rows were halved, now
presenting one row per ID for the 316 study subjects.

Then we tidyed the variables, by making new, more intuitive names fore
thoose laking of one. Columns containing information concerning two
variables were either separated (Subject-\> Hospital and ID ) or splitt
through pivot wider as mentioned. Columns listed in the assignmet as not
of interest was deleted, as asked for in the task. Columns were then
sorted with ID first.

Further on we made new variables to filter and explore the datasett. For
instance PVol_High as categorical varible ( \>/\< 100 ml) 
The variables
unit_norm, ttr_num, TimeToRecurrence_days_new, were all made in the
purpose of making the variable TimeToRecurrence_weeks_new, for
calculations with the same numeric category (weeks). We met a little
rocky road making this new variable, therefor keeping some duplicates to
remember the steps of the string.

Our datasett was then joined by the additional exam_data_joint.txt, by
the common variable ID. By the end of the tidying session we ended up
with 316 IDs( rows) and 27 variables (columns)

The coding was stored by the end of each session, with a script for the
same date. glimpse(tidy_exam_data_2025-09-09)
glimpse(tidy_exam_data-2025-09-08)
glimpse(joint_exam_data_data-2025-09-08) Please view
Updated_codebook_2025_11_09 and script for more details.

```{r cars}
library(here)
source(here("scripts", "tidy_exam_data_2025-09-08"))
source(here("scripts", "tidy_exam_data_2025-09-09"))
source(here("scripts", "joint_exam_data_data-2025-09-09"))



```

## stratifying the data

Explore your data. - Explore and comment on the missing variables. -
Stratify your data by a categorical column and report min, max, mean and
sd of a numeric column. - Stratify your data by a categorical column and
report min, max, mean and sd of a numeric column for a defined set of
observations - use a pipe `%>%`!\
*(each member of the group chooses one column)* - Only for persons with
`T.Stage == 1` - Only for persons with `Median.RBC.Age == 25` - Only for
persons with `TimeToReccurence` later than 4 weeks - Only for persons
recruited in `Hosp1` and `Tvol == 2` - Use two categorical columns in
your dataset to create a table (either with `table()`, `count()`, or
`janitor::tabyl()`)

list.files("scripts")







```{r}

```

## Including Plots

-   Are there any correlated measurements? (hint: `GGally::ggcorr` or
    search online for correlation matrix in R)
    -   Is there a relation between the `PVol` and `TVol` variables?
    -   Does the distribution of `PreopPSA` differ by `T.Stage`?
    -   Does the distribution of `PVol` differ by `sGS`?
    -   Does the distribution of `TVol` differ by `sGS`?
    -   Where there more `T.Stage == 2` in the group with
        `PreopTherapy == 1` than in the group `PreopTherapy == 0`?

You can also embed plots, for example:

```{r pressure, echo=FALSE}
source(here("scripts", "Script_ggplot_10-09-25.R"))

plot1
```

#Analysing the data\
After tidying, exploring, and making plots, we divided "task 5"
exercises among the group, which regarded exercises on analyzing the
data. Methods: Checked normality of time to recurrence (weeks)
(Shapiro–Wilk + histogram) → non-normal → used Kruskal–Wallis for group
comparisons. Furthermore, we compared time to recurrence across RBC
storage age groups and tumor stage. Then we ran two simple linear
models: TimeToRecurrence_days_new \~ Adjuvant_radiation_therapy; & TVol
\~ Recurrence. (Effect sizes and p-values reported in script output
tables.)

Results:RBC storage age groups: Kruskal–Wallis χ²(2)=1.022, p=0.60 → no
evidence of differences. Tumor stage: Kruskal–Wallis χ²(1)=7.962,
p=0.0048 → distributions differ by stage. Linear models showed:
Therapy–time and Recurrence–TVol associations summarized by
coefficients; see script tables for estimates and significance.

Conclusions: No difference by RBC storage age; clear difference by tumor
stage; AdjRadTherapy and time to recurrence showed no significant
association; Recurrence and TVol: Yes—patients with recurrence had
significantly larger TVol.

We also did a minor change in the dataset trough renaming the sGS column
to Surgical_gleason_score for tidying purposes.

See the full analysis script here:
[Script_sithday_2025-11-09.R](blank//scripts/Script_sithday_2025-11-09.R)

```{r include=FALSE}
library(here)
source(here("scripts", "Script_sithday_2025-11-09.R"))
```

Note that the `echo = FALSE` parameter was added to the code chunk to
prevent printing of the R code that generated the plot.
